### 2014-10-21 ApTranscriptome_TR.Rmd
modifing figures for manuscript

@@

### Comparison of the degree of upregulation against constitutive expression

To test the genetic assimilation hypothesis, I compare the degree of upregulation of a transcript against constitutive expression at optimal temperatures.

@@

# merge optimum expression, in Ap.response.type data.frame, with predicted TPM data
Ap.induction.wide <- data.table(Ap.induction.wide)
setkey(Ap.induction.wide, Transcript)
Ap.response.type2 <- Ap.induction.wide[Ap.response.type]

# plot constitutive expression vs induction for High, Low and Bimodal genes only
setkey(Ap.response.type2, A22.type)
up.terms <- c("Bimodal", "Low", "High")
Ap.response.type2.up <- Ap.response.type2[A22.type %in% up.terms, ]

setkey(Ap.response.type2.up, Ar.type)
Ap.response.type2.up <- Ap.response.type2.up[Ar.type %in% up.terms, ]
Ap.response.type2.up <- droplevels(Ap.response.type2.up)

gg5 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.upreg), colour = A22.type)) + 
  geom_point() + 
  geom_smooth() +
  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
  theme(legend.position = "none")

gg6 <- ggplot(Ap.response.type2.up, aes(x = log(A22.opt), y = log(Ar.upreg), colour = Ar.type)) + 
  geom_point() + 
  geom_smooth() +
  labs(x="log(consitutive expression of A. picea)", y = "log(Degree of inducibility in A. carolinensis)") +
  theme(legend.justification=c(1,0), legend.position=c(1,0), legend.text = element_text(size = 12))

# grid.arrange
png("results/constitutive_vs_inducibility.png")
grid.arrange(gg5, gg6, nrow = 1)
dev.off()

### plot constitutive expression in one species vs the other
gg7 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.opt), y = log(A22.opt), colour = A22.type)) + 
  geom_point() + 
  geom_smooth() +
  labs(x="log(consitutive expression of A. carolinensis)", y = "log(constitutive expression in A. picea)") +
  theme(legend.position = "none")  
gg7 <- gg7 + geom_abline(intercept=0, slope=1)

png("results/constitutive_comparison.png")
gg7
dev.off()

### plot inducibility in one species vs the other
gg8 <- ggplot(Ap.response.type2.up, aes(x = log(Ar.upreg), y = log(A22.upreg), colour = A22.type)) + 
  geom_point() + 
  geom_smooth() +
  labs(x="log(upregulation in A. carolinensis)", y = "log(upregulation in A. picea)") +
  theme(legend.position = "none")  
gg8

png("results/upregulation_comparison.png")
gg8
dev.off()

### constitutive vs upregulated for *High* genes only
Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
str(Ap.response.type2.A22.high)

gg9 <- ggplot(Ap.response.type2.A22.high, aes(x = log(Ar.opt), y = log(A22.opt), colour = Ar.type)) + 
  geom_point() + 
  geom_smooth() +
  labs(x="log(consitutive expression of A. carolinensis)", y = "log(Degree of inducibility in A. picea)") +
  theme(legend.position = "none")  
gg9 + geom_abline(intercept=0, slope=1)

png("results/constitutive_vs_inducibility_A22high.png")
gg9
dev.off()
```

>>>>

# make subplot of rows that match hypotheses

# A. carolinensis **Low** and **Bimodal**
md_sub1 <- subset(md, Acar %in% c("Low", "Bimodal"), )
md_sub1 <- droplevels(md_sub1)

mh_plot_sub1 <- qplot(x=Apic, y=Acar, data=md_sub1, fill=value, geom="tile", ylim = rev(levels(md_sub1$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") +
  theme(legend.position = "none")
mh_plot_sub1

md_sub2 <- subset(md, Apic %in% c("High", "Bimodal"), )
md_sub2 <- droplevels(md_sub2)

mh_plot_sub2 <- qplot(x=Apic, y=Acar, data=md_sub2, fill=value, geom="tile", ylim = rev(levels(md_sub2$Acar))) + 
  scale_fill_gradient2(limits=c(-600, 600)) +
  theme(axis.title = element_text(face="italic")) +
  labs(x = "A. picea", y = "A. carolinensis") + 
  theme(legend.position = "none")
mh_plot_sub2

# get legend from http://stackoverflow.com/questions/11883844/inserting-a-table-under-the-legend-in-a-ggplot2-histogram
#Extract Legend
g_legend <- function(a.gplot){
  tmp <- ggplot_gtable(ggplot_build(a.gplot))
  leg <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
  legend <- tmp$grobs[[leg]]
  return(legend)}

legend <- g_legend(mh_plot)

# remove legend from mh_plot
mh_plot <- mh_plot + theme(legend.position = "none")

# add bars
geom_rect(data = dark.subset, aes(xmin = phase_start, xmax = phase_end), 
              ymin = -Inf,
              xmax = Inf,
              fill = alpha("#000000", 0.2))

# arrange plots

grid.arrange(mh_plot, mh_plot_sub2, mh_plot_sub1, legend,
             ncol=2, widths=c(1,0.5), heights=c(1, 0.5))

arrangeGrob(mh_plot, mh_plot_sub1, mh_plot_sub2, ncol=1), 
                    ncol=2, widths=c(1,1.2))

@@

To visualize this data, I plot the log ratio of inducibility between species against the log ratio of baseline expression between species. I do this separately for **High** transcripts in *A. picea* and then **Low** transcripts in *A. carolinensis*.

First, I calculate the degree of expression induction as the proportion difference between the maximum and minimum expression levels for each transcript.

@@

# merge optimum expression, in Ap.response.type data.frame, with predicted TPM data
Ap.induction.wide <- data.table(Ap.induction.wide)
setkey(Ap.induction.wide, Transcript)
Ap.response.type2 <- Ap.induction.wide[Ap.response.type]
```

I extract the **High** transcripts for *A. picea* and plot the log ratio of inducibility in *A. picea* over inducibility in *A. carolinensis*, against the log ratio of baseline expression in *A. carolinensis* against expression in *A. picea*. According to the *genetic assimilation hypothesis* there should be a positive relationship between these ratios. That is, greater baseline expression in *A. carolinensis* should have greater inducibility towards **High** temperatures in *A. picea*.

```{r log_inducibility}
Ap.response.type2.A22.high <- Ap.response.type2[Ap.response.type2$Transcript %in% A22.high.transcripts.df$Transcript, ]
str(Ap.response.type2.A22.high)

# filter out NotResp transcripts in Ar
Ap.response.type2.A22.high <- Ap.response.type2.A22.high[which(Ap.response.type2.A22.high$Ar.type != "NotResp"), ]

# ratio of upregulation
Ap.response.type2.A22.high$A22.Ar.upreg <- Ap.response.type2.A22.high$A22.upreg / Ap.response.type2.A22.high$Ar.upreg

# ratio of baseline expression
Ap.response.type2.A22.high$Ar.A22.baseline <- Ap.response.type2.A22.high$Ar.opt / Ap.response.type2.A22.high$A22.opt

# plot!
gg11 <- ggplot(Ap.response.type2.A22.high, aes(y = log(A22.Ar.upreg+0.01), x = log(Ar.A22.baseline))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. picea / A. carolinensis"), ")")),
       x = expression(paste("log(baseline ", italic("A. carolinensis / A. picea"), ")")))    
gg11

A22.high.lm <- lm(log(A22.Ar.upreg) ~ log(Ar.A22.baseline), data = Ap.response.type2.A22.high)

summary(A22.high.lm)
```

In contrast to the prediction there is a negative relationship between these ratios. That is, genes with greater upregulation of expression also have greater baseline expression. This is also consistent with the results simply comparing baseline expression between species for the *A. picea* **High** transcripts.

I repeated the above but for **Low** transcripts in *A. carolinensis*.

```{r assimilation_Ar_low}
### Repeat for A. carolinensis *Low* transcripts

Ap.response.type2.Ar.low <- Ap.response.type2[Ap.response.type2$Transcript %in% Ar.low.transcripts.df$Transcript, ]
str(Ap.response.type2.Ar.low)

# filter out NotResp transcripts in A22
Ap.response.type2.Ar.low <- Ap.response.type2.Ar.low[which(Ap.response.type2.Ar.low$A22.type != "NotResp"), ]

# ratio of upregulation
Ap.response.type2.Ar.low$Ar.A22.upreg <- Ap.response.type2.Ar.low$Ar.upreg / Ap.response.type2.Ar.low$A22.upreg

# ratio of baseline expression
Ap.response.type2.Ar.low$A22.Ar.baseline <- Ap.response.type2.Ar.low$A22.opt / Ap.response.type2.Ar.low$Ar.opt

gg12 <- ggplot(Ap.response.type2.Ar.low, aes(y = log(Ar.A22.upreg), x = log(A22.Ar.baseline))) + 
  geom_point() + 
  geom_smooth(method = "lm") +
  labs(y = expression(paste("log(inducibility ", italic("A. carolinensis / A. picea"), ")")),
       x = expression(paste("log(baseline ", italic("A. picea / A. carolinensis"), ")")))

gg12

Ar.low.lm <- lm(log(Ar.A22.upreg) ~ log(A22.Ar.baseline), data = Ap.response.type2.Ar.low)

summary(Ar.low.lm)
```

I combine the two above figures into a single figure.

```{r fig3}
# grid.arrange
png(file = "results/genetic_assimilation.png")
grid.arrange(gg11, gg12, nrow = 1)
dev.off()
```

As with the previous comparison, I also find no evidence for the genetic assimilation hypothesis.
